$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-04 paper/paper.tex
syntax and semantics

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Syntax}
\label{syntax}
Let $x, y,\ldots$ range over \emph{variables} and $P, Q,\ldots$ over processes.
The syntax of the \picalc{} \cite{Sangio01} is given by grammar in \autoref{fig:syntax-names}.

\begin{figure}[h]
  \begin{subfigure}{1\textwidth}
    \begin{mathpar}
      \datatype
      {n : \N}
      {\Var_n : \Set}
      \; \textsc{Var}

@@

      \datatype
      {n : \N}
      {\Process_n : \Set}
      \; \textsc{Process}
    \end{mathpar}
  \end{subfigure}
  \begin{subfigure}[t]{.2\textwidth}
    \begin{equation*}
      \begin{split}
        P,Q ::=      & \; \PO                \\ 
                    |& \; (\new{}x) \; P       \\ 
                    |& \; \comp{P}{Q}        \\ 
                    |& \; \recvp{x}{y} P \\ 
                    |& \; \sendp{x}{y} P 
      \end{split}
    \end{equation*}
    \caption{Ill-scoped grammar with names}
    \label{fig:syntax-names}
  \end{subfigure}
  \hspace{.1\textwidth}
  \begin{subfigure}[t]{.7\textwidth}
    \begin{equation*}
      \begin{split}
        \Process_n ::=& \; \PO_n               &&\text{(process termination)}   \\ 
        |& \; \new{} \Process_{\suc n}         &&\text{(restriction)}          \\ 
        |& \; \comp{\Process_n}{\Process_n}    &&\text{(parallel composition)} \\ 
        |& \; \recv{\Var_n}{}\Process_{\suc n} &&\text{(input)}                \\ 
        |& \; \send{\Var_n}{\Var_n}\Process_n  &&\text{(output)}                  
      \end{split}
    \end{equation*}
    \caption{Well-scoped grammar with type-level de Bruijn indices}
    \label{fig:syntax}
  \end{subfigure}
  \caption{}
\end{figure}

Process $\PO$ denotes the terminated process, where no operations can further occur.
Process $(\new{}x)\; P$ creates a new channel $x$ bound with scope $P$.
Process $\comp{P}{Q}$ is the parallel composition of processes $P$ and $Q$.
Processes $\recvp{x}{y} P$ and $\sendp{x}{y} P$ denote respectively the input and output processes of a variable $y$ over a channel $x$, with continuation $P$.

Scope restriction $(\new{}x) \; P$ and input $\recvp{x}{y} \; P$ both introduce bound names in $P$.
In order to mechanise the \picalc{} syntax in Agda, we need to deal with bound names in continuation processes.
Names are cumbersome to mechanise: inserting a new variable into a context means proving that its name differs from all other variable names in context.
Furthermore, names do not guarantee that the syntax well-scoped.
Instead, we use de Bruijn indices \cite{deBruijn1972}, where a natural number $n$ (aka \emph{index}) is used to refer to the variable introduced $n$ binders ago --- hence binders no longer introduce names.
\begin{example}
  \begin{equation*}
    \begin{split}
      & (\new{}x ) && (\comp {\recvp{x}{y} && \; \sendp{y}{a} \; \PO} {(\new{} y) && (\sendp{x}{y} \; \recvp{y}{z} && \; \PO)})
      && \text{names}
      \\
      & \new{} && (\comp {\recv{0} && \; \send{0}{a} \; \PO} {\new{} && (\send{1}{0} \; \recv{0} && \; \PO)})
      && \text{de Bruijn}
      \\
    \end{split}
  \end{equation*}
\end{example}
That is, terms at different \emph{depths} use different indices to refer to the same binding.

A variable occurring under $n$ abstractions has $n$ possible variables to refer to.
References outside of that range are meaningless.
It is useful to rule out these nonsensical terms syntactically.
In \autoref{fig:syntax}, we do so by introducing the indexed family of types $\Var_n$: for all naturals $n$, the type $\Var_n$ has $n$ distinct elements.

Every time we go under a binder, the number of binders a variable might refer to increments by one.
To propagate this information, we index processes according to their \emph{depth}: for all naturals $n$, a process of type $\Process_n$ contains variables that can refer to $n$ distinct elements.
We increase the \emph{depth} counter by 1 in channel restriction and input processes.
\autoref{fig:syntax} showcases a \picalc{} syntax that is well-scoped by construction using type-level de Bruijn indices.

The semantics of our language can now be defined on the totality of the syntax.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Semantics}
\label{semantics}
We introduce the operational semantics for the \picalc{}, starting with structural congruence.

@@

The operational semantics of the \picalc{} is defined as a reduction relation \textsc{Reduces} indexed by the channel $i$ on which communication occurs (\autoref{fig:reduction}).
In rule $\constr{comm}$, parallel processes reduce when they communicate over a common channel ${i}$.
As a result of that communication, the continuation of the input process $P$ gets all the references to its most immediate variable substituted with references to $\suc j$, the variable sent by the output process $\send{i}{j}{Q}$.
After substitution, $\subst{P}{\suc j}{\constr{0}}$ is \emph{lowered} --- all variable references are decreased by one.
To do so, we apply \autoref{lm:subst-unused} in \autoref{app:lemmas} to obtain a proof $\Unused_{\constr{0}} \; (\subst{P}{\suc j}{\constr{0}})$.
Reduction is closed under parallel composition (rule $\constr{par}$), restriction (rule $\constr{res}$) and structural congruence (rule $\constr{struct}$) 
--- notably, not under input nor output, as doing so would not preserve the sequencing of actions \cite{Sangio01}.

We keep track of the channel over which communication occurs so that we can state subject reduction.
We use a $\func{dec}$ function to decrement the index of this channel every time we come out of a binder.
This function saturates at $\constr{internal}$: wrapping an internally reducing process with a binder will result in an internally reducing process.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Syntax}
\label{syntax}
\begin{figure}[h]
  \begin{subfigure}{1\textwidth}
    \begin{mathpar}
      \datatype
      {n : \N}
      {\Var_n : \Set}
      \; \textsc{Var}

@@

      \datatype
      {n : \N}
      {\Process_n : \Set}
      \; \textsc{Process}
    \end{mathpar}
  \end{subfigure}
  \begin{subfigure}[t]{.3\textwidth}
    \begin{equation*}
      \begin{split}
        P,Q ::=      & \; \PO                \\ 
                    |& \; (\new{}x) \; P       \\ 
                    |& \; \comp{P}{Q}        \\ 
                    |& \; \recvp{x}{y} P \\ 
                    |& \; \sendp{x}{y} P 
      \end{split}
    \end{equation*}
    \caption{Ill-scoped grammar with channel names}
    \label{fig:syntax-names}
  \end{subfigure}
  \hspace{.05\textwidth}
  \begin{subfigure}[t]{.65\textwidth}
    \begin{equation*}
      \begin{split}
        \Process_n ::=& \; \PO_n               &&\text{(inaction)}   \\ 
        |& \; \new{} \Process_{\suc n}         &&\text{(restriction)}          \\ 
        |& \; \comp{\Process_n}{\Process_n}    &&\text{(parallel)} \\ 
        |& \; \recv{\Var_n}{}\Process_{\suc n} &&\text{(input)}                \\ 
        |& \; \send{\Var_n}{\Var_n}\Process_n  &&\text{(output)}                  
      \end{split}
    \end{equation*}
    \caption{Well-scoped grammar with type-level de Bruijn indices}
    \label{fig:syntax}
  \end{subfigure}
  \caption{}
\end{figure}

Let $x, y,\ldots$ range over \emph{channel names} or \emph{variables} and $P, Q,\ldots$ over processes.
The syntax of the \picalc{} \cite{Sangio01} is given by the grammar in \autoref{fig:syntax-names}.
Process $\PO$ denotes the terminated process, where no operations can further occur.
Process $(\new{}x)\; P$ creates a new channel $x$ bound with scope $P$.
Process $\comp{P}{Q}$ is the parallel composition of processes $P$ and $Q$.
Processes $\recvp{x}{y} P$ and $\sendp{x}{y} P$ denote respectively the input and output processes of a variable $y$ over a channel $x$, with continuation $P$.

Scope restriction $(\new{}x) \; P$ and input $\recvp{x}{y} \; P$ both introduce bound names, respectively, $x$ and $y$ in $P$.
In order to mechanise the \picalc{} syntax in Agda, we need to deal with bound names in continuation processes.
Names are cumbersome to mechanise: inserting a new variable into a context means proving that its name differs from all other variable names in context.
Furthermore, names do not guarantee that the syntax is well-scoped.
\todo{It is not clear what ill vs well scoped means. Add a sentece saying: "By ill-scoped we mean... By well-scoped we mean... "}
Instead, we use de Bruijn indices \cite{deBruijn1972}, where a natural number $n$ (aka \emph{index}) is used to refer to the variable introduced $n$ binders ago --- hence binders no longer introduce names.
\begin{example}
  \begin{equation*}
    \begin{split}
      & (\new{}x ) && (\comp {\recvp{x}{y} && \; \sendp{y}{a} \; \PO} {(\new{} y) && (\sendp{x}{y} \; \recvp{y}{z} && \; \PO)})
      && \text{channel names}
      \\
      & \new{} && (\comp {\recv{0} && \; \send{0}{a} \; \PO} {\new{} && (\send{1}{0} \; \recv{0} && \; \PO)})
      && \text{de Bruijn indices}
      \\
    \end{split}
  \end{equation*}
\end{example}
That is, terms at different \emph{depths} use different indices to refer to the same binding introduced either by a restriction $\constr{\nu}$ or input $\constr{()}$.

A variable occurring under $n$ abstractions has $n$ possible variables to refer to.
References outside of that range are meaningless.
It is useful to rule out these nonsensical terms syntactically.
In \autoref{fig:syntax} (including the top), we do so by introducing the indexed family of types $\Var_n$: for all naturals $n$, the type $\Var_n$ has $n$ distinct elements.
Every time we go under a binder, the number of binders a variable might refer to increments by one, in rules $\constr{0}$ and $\suc n$. \todo{check this is right what i just said}
To propagate this information, we index processes according to their \emph{depth}: for all naturals $n$, a process of type $\Process_n$ contains variables that can refer to $n$ distinct elements.
We increase the \emph{depth} counter by 1 in channel restriction and input processes.
\autoref{fig:syntax} showcases a \picalc{} syntax that is well-scoped by construction using type-level de Bruijn indices.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Semantics}
\label{semantics}
Due to a well-scoped grammar in \autoref{fig:syntax}, the semantics of our language can now be defined on the totality of the syntax. We start by defining the structural congruence.

@@

The operational semantics of the \picalc{} is defined as a reduction relation \textsc{Reduces} indexed by the channel $i$ on which communication occurs (\autoref{fig:reduction}).
In rule $\constr{comm}$, parallel processes reduce when they communicate over a common channel ${i}$.
As a result of that communication, the continuation of the input process $P$ gets all the references to its most immediate variable substituted with references to $\suc j$, the variable sent by the output process $\send{i}{j}{Q}$.
After substitution, $\subst{P}{\suc j}{\constr{0}}$ is \emph{lowered} --- all variable references are decreased by one (to do so, we apply \autoref{lm:subst-unused} in \autoref{app:lemmas} to obtain a proof $\Unused_{\constr{0}} \; (\subst{P}{\suc j}{\constr{0}})$).
Reduction is closed under parallel composition (rule $\constr{par}$), restriction (rule $\constr{res}$) and structural congruence (rule $\constr{struct}$) 
--- notably, not under input nor output, as doing so would not preserve the sequencing of actions \cite{Sangio01}.

We keep track of the channel over which communication occurs so that we can state subject reduction (in \autoref{subject-reduction}).
We use a $\func{dec}$ function to decrement the index of this channel every time we come out of a binder.
This function saturates at $\constr{internal}$: wrapping an internally reducing process with a binder will result in an internally reducing process.
